Exploring the linear separablity of syntactic and semantic information in BERT embeddings

### 2022-12-12 paper/paper.tex
1st draft of added layerwise section

\begin{abstract}
%Motivating question and or topics of interest.
Syntax and semantics are at the foundation of every language, yet distinctions between the two are not readily agreed upon. We seek to explore how representations of these two information sets manifest in BERT embeddings. Specifically we investigate the degree of the linear separability of syntactic and semantic information in BERT embeddings, as well as quantify how important the linear component corresponding to one information set is to solving a classification task that targets the other information set. We use Iterative Nullspace Projection to decompose word-level BERT embeddings into syntactic, non-syntactic, semantic and non-semantic components to be used in syntactic and semantic classification tasks. Our results show that there is significant overlap between the syntactic and semantic components in BERT embeddings and the loss of either information set will negatively impact the performance of a linear classifier trained for either of the classification objectives. Somewhat surprisingly, we find that the loss of semantic information has a much greater impact on the performance of the syntax classifier than the loss of syntactic information has on the semantic classifier. 
%How we address it and our results.
%Simplified conclusion. 
\end{abstract}

@@

\begin{table}[ht]
    \centering
    \begin{tabular}{p{4cm}p{3cm}}\hline
        \textbf{Expression} & \textbf{Description}\\ \hline 
        $v_{orig}$ & Original BERT embeddings  \\
        $v_{nosem} = P_{sem} v_{orig}$ & Gained after INLP with semantic task \\
        $v_{nosyn}= P_{syn} v_{orig}$ & Gained after INLP with syntactic task \\
        $v_{sem} = v_{orig}-v_{nosem}$ & Semantic representation \\
        $v_{syn} = v_{orig}-v_{nosyn}$ & Syntactic representation \\
        Rand($n$) & BERT embeddings with n random directions removed \\
        %$v_{syn-sem} = P_{sem} v_{syn}$ &Contains syntactic features only \\
        %$v_{sem-syn} = P_{syn} v_{sem}$ &Contains semantic features only \\
        % $v_{syn*sem} = \langle v_{syn}, v_{sem} \rangle \cdot \frac{v_{syn}}{\|v_{sem}\|}$ & Syntactic representations projected on semantic space\\
        % $v_{sem*syn} = \langle v_{syn}, v_{sem} \rangle \cdot \frac{v_{syn}}{\|v_{syn}\|}$ & Semantics representations projected on semantic space\\

@@

\section{Results}
\label{sec:result}

We first evaluate our two tasks on the original embeddings, and determine that linear classifiers can successfully predict both CCG tags and semantic tags (around 85\% testing accuracy), as shown in table \ref{role description}. We then apply the INLP method to derive the guarding matrices $P_{syn}$ and $P_{sem}$, which are used to project the original embeddings onto the complements of the syntactic information sub-space and the semantic information sub-space. By applying linear transformations to the original embeddings and their projections, we are able to extract the embeddings described in table \ref{description}.

To ensure a fair assessment of the impact of the information loss, we conduct experiments for which we start with the original BERT embeddings and randomly remove the same number of directions that our derived embeddings lost. For the embeddings derived from projection matrices, we record the number of removed directions using $Null(P)$, where $P$ is the projection matrix. .\footnote{For embeddings not derived from matrix multiplication, we obtain the number from $Null(M)$ where $M$ is the embedding matrix with size (768,instance number). This is not necessarily equivalent to the number of direction removed. $Null(M) \geq Null(P)$ for the embedding matrix $M$ that corresponds to projection $P$, but in practice the numbers are very close. With this approach we err on the side of removing more random directions from our random embeddings, resulting in a more conservative comparison when assessing our derived embeddings. Also note that we decided not to include these types of embeddings in our final assessment due to concerning behavior we witnessed after performing the linear transformation to obtain the embeddings. Specifically, the rank of the resulting embedding matrix did not match our expectations. This may be due to a rounding or an issue with the orthonormalization process, but we were unable to confirm. So we decided to remove all embeddings derived in this manner from our evaluation process until we are able to determine the source of this behavior. This means that $v_{syn}$ and $v_{sem}$ will not be included in our final evaluation.} Then we create embeddings with the same number of directions randomly removed and train the linear classifiers on these embeddings. The testing accuracies from our experiments can be found in table \ref{role description}.

@@

    Rand($|v_{nosem}|$) & 77 & $84.06\%$ & $87.71\%$ \\
    Rand($|v_{nosyn}|$) & 159 & $82.26\%$ & $87.57\%$ \\
    %Rand($|v_{syn-sem}|$) & 81 & $83.11\%$ & $87.61\%$ \\
    %Rand($|v_{sem-syn}|$) & 161 & $82.46\%$ & $86.04\%$ \\
    % Rand($|v_{syn*sem}|$) & 0 & p$\%$ & p$\%$ \\
    % Rand($|v_{sem*syn}|$) & 0 & p$\%$ & p$\%$ \\
    \hline
    $v_{nosem}$ & 77 & $27.93\%$ & $17.28\%$ \\
    $v_{nosyn}$ & 159 & $23.76\%$ & $49.43\%$ \\
    %$v_{syn-sem}$ & 81 & $34.24\%$ & $33.91\%$ \\
    %$v_{sem-syn}$ & 161 & $26.96\%$ & $50.21\%$ \\
    % $v_{syn*sem}$ & 0 & p$\%$ & p$\%$ \\
    % $v_{sem*syn}$ & 0 & p$\%$ & p$\%$ \\
    \hline
    \hline
    \end{tabular}
    \caption{\label{role description} Experiment Result of Different Embeddings
    }
    \end{table*}

@@

To measure the importance of semantic information for the syntactic task, we compare the performance of the CCG tagging classifier on $v_{nosem}$ with that of Rand($|v_{nosem}|$) and see a performance drop of 56.13\%. Similarly, we can measure the importance of syntactic information for the semantic task by comparing the performance of the semantic tagging classifier on $v_{nosyn}$ with that of Rand($|v_{nosyn}|$) and see a performance drop of 38.14\%. We can see that the removal of each information set has a significant impact on the performance of a linear classifier trained on either classification task. This suggests that the syntactic and semantic information in BERT embeddings is not easily disentangled.

@@

@@

@@

Though INLP successfully produces the desired result, it is worth noting that our dataset is relatively small compared to the number of parameters in the linear classifier. In the future, we will expect this experiment to be reproduced at a larger scale so that the experiment findings can be further validated. Additionally, we plan to increase the variety of training and evaluation tasks in order to develop a more general understanding of how syntactic and semantic information is encoded in BERT embeddings. Finally, we intend to perform a similar analysis on embeddings derived from different layers of the BERT architecture, in order to determine the importance and separability of these information sets at each layer. In doing so we hope to better understand how BERT processes different types of linguistic information throughout the encoding process.

>>>>

\begin{abstract}
%Motivating question and or topics of interest.
Syntax and semantics are at the foundation of every language, yet distinctions between the two are not readily agreed upon. We seek to explore how representations of these two information sets manifest in BERT embeddings. Specifically we investigate the degree of the linear separability of syntactic and semantic information in BERT embeddings, as well as quantify how important the linear component corresponding to one information set is to solving a classification task that targets the other information set. We use Iterative Nullspace Projection to decompose word-level BERT embeddings into syntactic, non-syntactic, semantic and non-semantic components to be used in syntactic and semantic classification tasks. Our results show that there is significant overlap between the syntactic and semantic components in BERT embeddings and the loss of either information set will negatively impact the performance of a linear classifier trained for either of the classification objectives. Somewhat surprisingly, we find that the loss of semantic information has a much greater impact on the performance of the syntax classifier than the loss of syntactic information has on the semantic classifier. Our results of layer-wise analysis also provide incentive to set an additional stopping condition of INLP, regarding the rank of the projection matrix.
%How we address it and our results.
%Simplified conclusion. 
\end{abstract}

@@

\subsection{Layer-wise evaluation}
In addition to the final layer BERT embeddings, we perform a similar analysis on the embeddings derived from different layers of the BERT architecture, in order to determine the separability of these information sets at each layer. For embedding $v_{orig_i}$ from layer $i,$ a linear classifier is trained for each probing task to acquire guarding functions $P_{syn_i}$ and $P_{sem_i},$ respectively. Applying these projection functions, we are able to acquire $v_{nosyn_i}$ and $v_{nosem_i}.$ Subtracting them from the original embedding, we get the semantic representation $v_{sem_i}$ and the syntatic representation $v_{syn_i}.$ We also randomly remove the same number of dimensions in the original embedding for comparison.

By comparing the experiment results across different information sets and different layers, we hope to better understand how BERT processes different types of linguistic information throughout the encoding process.

\begin{table}[ht]
    \centering
    \begin{tabular}{p{4cm}p{3cm}}\hline
        \textbf{Expression} & \textbf{Description}\\ \hline 
        $v_{orig}$ & Original BERT embeddings  \\
        $v_{nosem} = P_{sem} v_{orig}$ & Gained after INLP with semantic task \\
        $v_{nosyn}= P_{syn} v_{orig}$ & Gained after INLP with syntactic task \\
        $v_{sem} = v_{orig}-v_{nosem}$ & Semantic representation \\
        $v_{syn} = v_{orig}-v_{nosyn}$ & Syntactic representation \\
        Rand($v$, $n$) & embeddings $v$ with n random directions removed \\
        Subscript $i$ & Objects relating to the $i$-th layer of the BERT architecture rather than the final layer\\
        %$v_{syn-sem} = P_{sem} v_{syn}$ &Contains syntactic features only \\
        %$v_{sem-syn} = P_{syn} v_{sem}$ &Contains semantic features only \\
        % $v_{syn*sem} = \langle v_{syn}, v_{sem} \rangle \cdot \frac{v_{syn}}{\|v_{sem}\|}$ & Syntactic representations projected on semantic space\\
        % $v_{sem*syn} = \langle v_{syn}, v_{sem} \rangle \cdot \frac{v_{syn}}{\|v_{syn}\|}$ & Semantics representations projected on semantic space\\

@@

\section{Results}
\label{sec:result}
\subsection{Final layer evaluation}
We first evaluate our two tasks on the original embeddings, and determine that linear classifiers can successfully predict both CCG tags and semantic tags (around 85\% testing accuracy), as shown in table \ref{role description}. We then apply the INLP method to derive the guarding matrices $P_{syn}$ and $P_{sem}$, which are used to project the original embeddings onto the complements of the syntactic information sub-space and the semantic information sub-space. By applying linear transformations to the original embeddings and their projections, we are able to extract the embeddings described in table \ref{description}.

To ensure a fair assessment of the impact of the information loss, we conduct experiments for which we start with the original BERT embeddings and randomly remove the same number of directions that our derived embeddings lost. For the embeddings derived from projection matrices, we record the number of removed directions using $Null(P)$, where $P$ is the projection matrix. \footnote{For embeddings not derived from matrix multiplication, we obtain the number from $Null(M)$ where $M$ is the embedding matrix with size (768,instance number). This is not necessarily equivalent to the number of direction removed. $Null(M) \geq Null(P)$ for the embedding matrix $M$ that corresponds to projection $P$, but in practice the numbers are very close. With this approach we err on the side of removing more random directions from our random embeddings, resulting in a more conservative comparison when assessing our derived embeddings. Also note that we decided not to include these types of embeddings in our final assessment due to concerning behavior we witnessed after performing the linear transformation to obtain the embeddings. Specifically, the rank of the resulting embedding matrix did not match our expectations. This may be due to a rounding or an issue with the orthonormalization process, but we were unable to confirm. So we decided to remove all embeddings derived in this manner from our evaluation process until we are able to determine the source of this behavior. This means that $v_{syn}$ and $v_{sem}$ will not be included in our final evaluation.} Then we create embeddings with the same number of directions randomly removed and train the linear classifiers on these embeddings. The testing accuracies from our experiments can be found in table \ref{role description}.

@@

    Rand($v_orig, |v_{nosem}|$) & 77 & $84.06\%$ & $87.71\%$ \\
    Rand($|v_orig, v_{nosyn}|$) & 159 & $82.26\%$ & $87.57\%$ \\
    %Rand($|v_{syn-sem}|$) & 81 & $83.11\%$ & $87.61\%$ \\
    %Rand($|v_{sem-syn}|$) & 161 & $82.46\%$ & $86.04\%$ \\
    % Rand($|v_{syn*sem}|$) & 0 & p$\%$ & p$\%$ \\
    % Rand($|v_{sem*syn}|$) & 0 & p$\%$ & p$\%$ \\
    \hline
    $v_{nosem}$ & 77 & $27.93\%$ & $17.28\%$ \\
    $v_{nosyn}$ & 159 & $23.76\%$ & $49.43\%$ \\
    %$v_{syn-sem}$ & 81 & $34.24\%$ & $33.91\%$ \\
    %$v_{sem-syn}$ & 161 & $26.96\%$ & $50.21\%$ \\
    % $v_{syn*sem}$ & 0 & p$\%$ & p$\%$ \\
    % $v_{sem*syn}$ & 0 & p$\%$ & p$\%$ \\
    \hline
    \hline
    \end{tabular}
    \caption{\label{role description} Experiment Result of Different Embeddings
    }
    \end{table*}

\subsection{Intermediate Layer Evaluation}

Linear classifiers are generally able to achieve a test accuracy greater than 90\% for both CCG tagging and semantics tagging. However, after applying the INLP algorithm and projecting the layer-wise embeddings to the nullspaces of the information sets, we observe that we get the same results across all layers for each evaluation task, with the accuracy being the majority class of the evaluation task. Evaluations of Rand($v_i, |v_{nosyn_i}|$) and Rand($v_i, |v_{nosem_i}|$) result in the same majority class accuracy.

Upon a close inspection of the INLP process and the projections of the original intermediate layer embeddings, $v_{nosem_i}$ and $v_{nosyn_i}$, we realize that, the INLP process continues to run even if it already removes more ranks than BERT’s hidden size, which is 768 in our case, because the desired dev accuracy is still not met. Once the rank of the projection matrix reaches the limit, the INLP process simply reduces the magnitude of each elements in the embeddings. In some cases, the process eventually zeroes out the embeddings, which explains the identical yet trivial result we get from the evaluation task.

@@

To measure the importance of semantic information for the syntactic task, we compare the performance of the CCG tagging classifier on $v_{nosem}$ with that of Rand($v_orig, |v_{nosem}|$) and see a performance drop of 56.13\%. Similarly, we can measure the importance of syntactic information for the semantic task by comparing the performance of the semantic tagging classifier on $v_{nosyn}$ with that of Rand($v_{orig}, |v_{nosyn}|$) and see a performance drop of 38.14\%. We can see that the removal of each information set has a significant impact on the performance of a linear classifier trained on either classification task. This suggests that the syntactic and semantic information in BERT embeddings is not easily disentangled.

@@

As an unexpected finding of our experiment, we are unable to fully remove the syntactic/semantic information from the intermediate embedding by training the linear classifier to make prediction that is no better than the majority, without removing more ranks than BERT's hidden size, on all intermediate layers of BERT. However, removing more ranks than BERT’s hidden size, whether through the INLP algorithm or randomly, results in a degenerate embedding where every element is reduced to an extremely small magnitude that the linear probe on the evaluation task will only reach the majority class accuracy. This seems to reveal that, on the intermediate layers of BERT, the target information is not linearly separable from the original embedding. Besides using the majority class accuracy as the stopping condition, researchers hoping to use INLP to guard information from BERT embeddings should also make sure the loop stops before removing too many ranks for non-trivial results.

@@

By performing a similar analysis on embeddings derived from different layers of BERT architecture, we discovered that the syntactic information set and semantic information set are not linearly separable from the original intermediate layer embeddings. Attempting to completely remove these information sets will remove too much that the embeddings become degenerate.

@@

Though INLP successfully produces interesting results on various tasks, it is worth noting that our dataset is relatively small compared to the number of parameters in the linear classifier. Reproduing this experiment at a larger scale will be helpful in further validating the experiment results. Additionally, the variety of training and evaluation tasks can be increased for a broader understanding of how syntactic and semantic information is encoded in BERT embeddings.
